---
title: Testing differences in results and time between neighborhood analysis in R
  and GRASS
author: "Bernardo Niebuhr"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dpi = 300)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Here we'll test the differences between different approaches to make filters and calculate densities in both R and
GRASS GIS. In GRASS, we'll calculate densities using two different modules, `r.neighbors` and
`r.resamp.filter` to check the difference in their outputs, both qualitatively and in terms of time spent
in the calculations.

After that, we'll also perform neighborhood analysis in R using the `focal` function of the `terra` package and
compare the computation time with the GRASS GIS approaches.

We'll first look at the qualitative differences in a small region, then we perform the operations for the whole
Norway for comparing computation time.

## Set environment

I'll work in my personal mapset in the "ETRS_33N" GRASS GIS location, copying a map of forest amount from NORUT and
calculating densities based on it.

```{r settings, results="hide", message=FALSE, warning=FALSE}
# libraries
library(rgrass7)
library(raster)
library(terra)
library(NinaR)

library(dplyr)

library(microbenchmark)
library(tictoc)
library(viridis)
```

Connect to GRASS.

```{r connect2GRASS}
# should we run the heavy tests?
heavy_test <- TRUE

# start GRASS and connect to my mapset
grassdir <- system("grass78 --config path", intern = T)
gisDB <- "/data/grass"
loc <- "ETRS_33N/"
ms <- "u_bernardo.brandao"

# initGRASS(gisBase = grassdir,
#           home = tempdir(), 
#           override = T,
#           gisDbase = gisDB,
#           location = loc, 
#           mapset = ms)

# more directly within NINA
NinaR::grassConnect(mapset = ms)
```

```{r}
# system("export GRASS_MESSAGE_FORMAT=silent")
stringexecGRASS("g.gisenv set='GRASS_MESSAGE_FORMAT=silent'") 
```

Prepare the input maps in both GRASS and R and visualize them.

```{r prepare_input}
# copy map of forests from NORUT

# check region
gmeta() # g.copy is not affected by the region

# copy
# execGRASS("g.copy", flags = "help")
execGRASS("g.copy", parameters = list(raster = "norut_skog_100@p_RenRein_norut,norut_skog_100"),
          flags = c("overwrite"))

# define region
# I defined a region_test through the GRASS GUI and using v.in.region output=region_test
# execGRASS("g.copy", parameters = list(vector = "region_test@u_bernardo.brandao2,region_test"))
# then I removed the vector from the other mapset
execGRASS("g.region", parameters = list(vector = "region_test", align = "norut_skog_100"),
          flags = "p")

# create a new small map for the current region, just to ease visualization in R
execGRASS("r.mapcalc", expression = "norut_skog_100_sub = norut_skog_100",
          flags = c("overwrite", "quiet"))

# show input map and region here
use_sp()
forest <- readRAST(c("norut_skog_100_sub"))

# as a SpatialGridDataFrame
image(forest, "norut_skog_100_sub", col = viridis(20))

# raster
r <- raster::raster(forest)
# raster::plot(r)
# terra
r2 <- terra::rast(r)
plot(r2)
```

## Test1: within GRASS, the difference between `r.neighbors` and `r.resamp.filter`

Here we start by comparing qualitatively the results of the two GRASS GIS modules for the 
small test region..

```{r test1_grass, warning=FALSE, message = FALSE}
# r.resamp.filter
# execGRASS("r.resamp.filter", flags = "help")

tic()
execGRASS("r.resamp.filter",
          parameters = list(input = "norut_skog_100_sub", output = "test1_resamp_filter_500",
                            filter = "bartlett", radius = 500),
          flags = c("overwrite", "quiet"))
toc()

# r.neighbors
# execGRASS("r.neighbors", flags = "help")

tic()
execGRASS("r.neighbors",
          parameters = list(input = "norut_skog_100_sub", output = "test1_neighbors_500",
                            method = "average", size = 9),
          flags = c("c", "overwrite", "quiet"))
toc()

# show output
g1 <- readRAST(c("test1_resamp_filter_500", "test1_neighbors_500"))

# to SpatRaster
g1_1 <-  raster::raster(g1) %>% 
  raster::stack(raster(g1, layer = 2)) %>% 
  terra::rast()

plot(g1_1)
```

From the plot, the results seem quite similar. Indeed, the mean, median, and 1st/4th quantiles are close to each other. However, they differ locally in their values, differences of up to .1 (for a quantity in the interval [0, 1]), in this case with average `as.numeric(global(g1_1[[1]], "mean"))`. Most probably this is due to different algorithms for filtering and averaging.

```{r}
# Summary statistics
summary(g1_1)

# Plot difference
# plot(diff(g1_1)/g1_1[[1]], main = "Difference of outputs r.resamp.filter - r.neighbors")
plot(diff(g1_1), main = "Difference of outputs r.resamp.filter - r.neighbors")
```

## Test2: qualitative comparison of GRASS vs R neighborhood operations

Now we do the same including an R implementation for neighborhood analysis, with the `terra` package.

```{r}
# ?terra::focal

tic()
terra_focal <- focal(r2, w = 9, fun = "mean")
toc()
names(terra_focal) <- "focal_500"

# merge outputs
g2 <- c(g1_1, terra_focal)

# Plot
plot(g2)

# Summary statistics
summary(g2)

# Plot difference
# plot(diff(g1_1)/g1_1[[1]], main = "Difference of outputs r.resamp.filter - r.neighbors")
plot(diff(g2[[c(1,3)]]), main = "Difference of outputs r.resamp.filter - focal")
```

Again the outputs are similar, but there are local differences.

## Test3: computation time in GRASS vs R operations
 
Now we use the microbenchmark function to compare the computation time between each approach,
first for a small area, then using bigger window sizes, and finally for the whole Norway.

```{r}
# ?microbenchmark
# defining function bna = benchmark for neighborhood analysis
source("code/functions.R")

# Parameters
input_map_grass <- "norut_skog_100_sub"
input_map_r <- r2

# Window size
size_m = 500
size_pixels = 2*size_m/100 - 1#9

(bm1 <- bna(input_map_grass, input_map_r, size_m, size_pixels))
```

Here, interestingly, R is faster than GRASS.

Now we change the window size to something bigger - let's use 2500m (or 49 pixels).

```{r}
# Window size
size_m = 2500
size_pixels = 2*size_m/100 - 1

(bm2 <- bna(input_map_grass, input_map_r, size_m, size_pixels, times = 10))
```

Here the picture is different. `r.resamp.filter` is much faster, and `r.neighbors` and `focal` have comparable computation times. This keeps the picture in the tests below.

We can take even bigger window size - let's use 10,000m (or 199 pixels).

```{r}
# Window size
size_m = 10000
size_pixels = 2*size_m/100 - 1

if(heavy_test) (bm3 <- bna(input_map_grass, input_map_r, size_m, size_pixels, times = 10))
```

Now we run for the whole country, but only once, just to have an idea.
Let's keep an intermediate scale for the window size, 2500m, for that test.

```{r}
# Region
execGRASS("g.region", parameters = list(raster = "norut_skog_100"),
          flags = "p")

# Input as SpatRaster
skog <- readRAST("norut_skog_100") %>% 
  raster::raster() %>% 
  terra::rast()

# input map
input_map_grass <- "norut_skog_100"
input_map_r <- skog

# Window size
size_m = 2500
size_pixels = 2*size_m/100 - 1#9

if(heavy_test) (bm4 <- bna(input_map_grass, input_map_r, size_m, size_pixels, times = 1))
```

